\LaTeXe\ guide for the Journal of Functional Programming

### 2019-04-25 papers/jfp/8-engineering.tex
Expand S8.1: Partial stores and exceptions

@@

Our model assumes a world where the store is fully-defined, every \hs{k} is
associated with a \hs{v}, and every compute successfully completes returning a
valid value. In the real world, build systems frequently deal with errors, e.g.
``file not found'' or ``compilation failed''. We can model such failure
conditions by instantiating \hs{v} to either \hs{Maybe}~\hs{v} (for missing
values) or \hs{Either}~\hs{e}~\hs{v} (for exceptions of type \hs{e}). That can
model the errors inside the store and the \hs{Task}, but because \hs{v} is
polymorphic for builds, it does not let the build system apply special behaviour
for errors, e.g. early aborting.

>>>>

\todo{AM}{Tasks with multiple outputs:
  https://github.com/snowleopard/build/blob/master/src/Build/Multi.hs}

\todo{AM}{Typed tasks:
  https://github.com/snowleopard/build/blob/master/src/Build/Task/Typed.hs}

\todo{AM}{Self-tracking (?):
  https://github.com/snowleopard/build/blob/master/src/Build/SelfTracking.hs}

\todo{AM/NM}{Re-using existing infrastructure: (a) key-value store, (b) remote execution service.}

@@

Our model assumes a world where the store is fully-defined, every \hs{k} is
associated with a \hs{v}, and every compute successfully completes returning a
valid value. In the real world, build systems frequently deal with errors, e.g.
``file not found'' or ``compilation failed''. We can model such failure
conditions by instantiating \hs{v} to either \hs{Maybe}~\hs{v} (for missing
values) or \hs{Either}~\hs{e}~\hs{v} (for exceptions of type \hs{e}). That can
model the errors inside the store and the \hs{Task}, but because our build
system models are polymorphic in \hs{v}, they cannot apply any special behaviour
for errors, e.g. aborting the build early. There are many other ways of
modelling errors, and in this section we show three simple examples.

As mentioned above, we can include failures into the type of values \hs{v}, for
example, to model a partial store we can use a simple like algebraic data type
isomorphic to \hs{Maybe}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Value = FileNotFound | FileContents ByteString
\end{minted}
\vspace{1mm}

\noindent
This is convenient if \emph{tasks are aware of failures}. For example, a task
may be able to cope with missing files, e.g. if \hs{fetch}~\hs{"username.txt"}
returns \hs{FileNotFound}, the task could use the literal string \hs{"User"} as
a default value. In this case the task will \emph{depend} on the fact that the
file \cmd{username.txt} is missing, and will need to be rebuilt if the user
later creates this file. In general, we can use values of type
\hs{Either}~\hs{e}~\hs{v} when dealing with failures of type \hs{e}. To
automatically convert a ``failure-free'' task into an equivalent task operating
with values of type \hs{Either}~\hs{e}~\hs{v} we can use the following function:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
liftEither :: Task Monad k v -> Task Monad k (Either e v)
liftEither task = Task $ \fetch ->
    runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{1mm}

\noindent
Here \hs{liftEither} wraps the result of the given
\hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{Either}~\hs{e}~\hs{v} into \hs{EitherT}
and then runs the task in the \hs{EitherT} monad transformer~\cite{liang1995monad}.

Another approach is to include failures into the computation context \hs{f}. For
example, we can choose \hs{f}~\hs{=}~\hs{Maybe}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
fetchA1A2 :: String -> Maybe Integer
fetchA1A2 k = Map.lookup k (Map.fromList [("A1", 10), ("A2", 20)])
\end{minted}
\vspace{1mm}

\noindent
Since tasks are polymorphic in \hs{f}, we can directly run them with the
callback \hs{fetchA1A2}. For example, the task description \hs{sprsh1}
from~\S\ref{sec-task} will return \hs{Just}~\hs{(Just}~\hs{30)} for the task
\cmd{B1 = A1 + A2}, \hs{Just}~\hs{Nothing} for the task \cmd{B2 = B1 * 2}
because \hs{fetchA1A2} failed, and simply \hs{Nothing} for the input \cmd{A1}.

This approach is convenient if \emph{tasks are not aware of failures}, e.g. we
can model Excel formulas as pure arithmetic functions, and introduce failures
``for free'' if/when needed by instantiating \hs{Tasks} with an appropriate
\hs{f}.

Finally, the task itself might not want to encode failures into the type of
values \hs{v}, but instead \emph{demand that} \hs{f} \emph{has a built-in notion
of failures}. This can be done by choosing a suitable constraint \hs{c}, such as
\hs{Alternative}, \hs{MonadPlus} or even better something specific to failures,
such as \hs{MonadFail}. Then both the callback and the task can reuse the same
failure mechanism as shown below:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
class Monad m => MonadFail m where
    fail :: String -> m a

sprsh4 :: Tasks MonadFail String Integer
sprsh4 "B1" = Just $ Task $ \fetch -> do
    a1 <- fetch "A1"
    a2 <- fetch "A2"
    if a2 == 0 then fail "division by 0" else return (a1 `div` a2)
sprsh4 _ = Nothing
\end{minted}
\vspace{1mm}

\noindent
With this approach we can define a \hs{MonadFail} build system that can handle
errors by aborting the build early.

\todo{AM}{Add a MonadFail build system.}
